Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-11 conference-materials/2025_posit-conf/draft.qmd
Refactor slide content for clarity and narrative flow
- Promoted "Core Realization" into a dedicated slide to emphasize its importance
- Removed redundant inline styling (`margin: 5% 10%`) to improve layout consistency
- Restructured behavioral science and paradox of choice content into clearer sections
- Applied consistent `.fragment` class usage to improve reveal timing in slides
- Enhanced note sections with more precise speaker guidance

:::: {.columns .fragment}
::: {.column width="45%"}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="50%"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%"}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="60%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

@@

<div style="margin: 5% 10%;">

:::: {.columns}
::: {.column width="50%" .text-xl}
### Raise hand if you've ever:
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?

@@

::: {.column width="50%" .fragment .text-xl}
### 👋 **Hi, I'm Jeremy!

**PhD in Applied Psychology**<br>
*How people actually process information*

**Led Shiny teams at scale**<br>
*Built dashboards people use*

**Created {bidux} package**<br>
*Your behavioral scientist in the console*

I'm a psychologist by training — but not the kind you're thinking of.**<br>

*Contrary to popular belief, most psychologists work with data, not clients.*<br>

**Academic → Tech trajectory**: Applied psychology PhD → Full stack engineer
:::
::::

::: {.fragment}
### Core Realization

> "Most dashboard failures aren't technical. They're psychological."
:::

</div>

::: {.notes}
- 1 minute
- Quick show of hands...who's created a user interface with more than 10 filters?
- Or if ever had a client/user ask for "just one more dropdown"
- Or built a Shiny app with more `selectInput()`s than insights
- [CLICK] If so...welcome you to my dashboard support group
- Think of our time together today as a little therapy session for dropdown overload
- [PAUSE] Hi, I'm Jeremy! I'm a psychologist by training...but not the kind you're thinking of!
- Contrary to popular belief...there are many types of psychologists...and most of us with data, not clients!
- Clarify the therapy metaphor: support group hook connects our shared pain to therapeutic framing
- Most psychologists actually work with data, studying behavior patterns, not doing individual therapy
- My path: academic research → tech industry, bringing behavioral insights to engineering challenges
- "PhD in Applied Psych...studied how people/groups process information and make decisions"
- Transitioned from academia to industry...where I got involved leading Shiny teams at scale
- "Have nearly 10 years experience building enterprise apps/dashboards folks actually use"
- But it wasn't always that way...I've built MANY dashboards that flopped even though they were technically sound
- This is what led to my core realization: Most dashboard failures aren't technical, they're psychological
- "Let me show you what I mean"
:::

## The Problem is Real {background-image="img/ui-before.png" background-size="contain" background-position="right" background-color="#fff"}

::: {.absolute left="-50" bottom="200"}
### Real client. Real dashboard. Real failure.

@@

<br>

::: {.fragment}
**We ran telemetry for 30 days...**
:::
:::

::: {.notes}
- 45 seconds
- "I worked with a client who had this 'comprehensive' dashboard"
- Blinded here for confidentiality...but incredibly close to the actual design
- "18 filters, 6 tabs, every possible view, nearly zero adoption"
- Why this matters: This represents thousands of developer hours and client budget wasted on technically perfect but psychologically broken software
- The client was frustrated bc they had spent months building this thing, but no one used it
- As developers, we can relate - we've all built features nobody uses. But here's why this matters more than frustration:
    1. Technical debt from unused complexity compounds over time
    2. User abandonment patterns replicate across similar projects
    3. Without data-driven insights, we keep making the same UX mistakes
- [CLICK] "So we ran telemetry for 30 days..." - This data changed everything about how I approach dashboard design
:::

## The Telemetry Revealed Everything

@@

ggplot(telemetry_data, aes(x = metric, y = percentage)) +
  geom_col(
    fill = c("#27ae60", "#e74c3c", "#c0392b", "#a93226", "#e74c3c"),
    width = 0.7
  ) +
  geom_text(
    aes(label = paste0(percentage, "%")),
    vjust = -0.5,
    size = 10,
    fontface = "bold",
    # color = "white"
  ) +
  geom_segment(
    aes(
      x = as.numeric(metric) - 0.35,
      xend = as.numeric(metric) + 0.35,
      y = percentage,
      yend = percentage
    ),
    # color = "white",
    size = 1,
    alpha = 0.3
  ) +
  labs(
    title = "The User Journey of Despair",
    subtitle = "1,847 sessions tracked over 30 days",
    x = "",
    y = "Percentage of Users"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    # plot.background = element_rect(fill = "#e74c3c", color = NA),
    # panel.background = element_rect(fill = "#e74c3c", color = NA),
    plot.title = element_text(size = 36, face = "bold"),
    plot.subtitle = element_text(size = 24),
    axis.text.x = element_text(
      size = 16,
      angle = 25,
      hjust = 1
    ),
    axis.text.y = element_text(size = 18),
    # axis.title = element_text(color = "white"),
    panel.grid = element_blank()
  ) +
  ylim(0, 110)
```

@@

::: {.notes}
- 45 seconds
- "Look at this journey of despair"
- "100% open, only 35% click anything"
- Only 8% succeed in finding what they needed...and we didn't collected data on this..but I would hypothesize that many of those folks were a bit more data savvy and just knew where to look
- [CLICK] Overall...this was essentially 92% failure rate on a product that cost the client thousands of dollars to build
- This is behavioral science screaming at us that something is wrong here
:::

@@

<div style="margin: 5% 10%;">

@@

::: {.fragment}
### Applied to Dashboards

- **Cognitive Load**: Too many choices
- **Progressive Disclosure**: Reveal complexity gradually
- **Framing Effects**: Same data, different story
- **Choice Architecture**: Guide without restricting
:::

</div>

::: {.notes}
- 30 seconds
- "Behavioral science studies how people actually make decisions"
- "Irrationally but predictably"
- [CLICK] "Applied to dashboards"
- "These aren't just theories, they're tools"
:::

## Classic Example: The Paradox of Choice {background-color="#5b6bbf"}

:::: {.columns}
::: {.column width="50%"}
![Jam Study Results](img/jam-study-behavioral.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::

@@

::: {.notes}
- 45 seconds
- "Classic behavioral science example that perfectly mirrors our dashboard problem"
- "Sheena Iyengar's famous jam study at Columbia University"
- Show image: Side-by-side grocery store displays - one with 24 jam varieties (crowded, overwhelming), one with 6 varieties (clean, organized). Include simple bar charts showing the dramatic difference in purchase rates.
- "More choice attracted browsers, but paralyzed buyers"
- [CLICK] "This is exactly what happens with dashboard filters - more options kill engagement"
- "This principle drives everything we'll build with the BID framework"
:::

@@

</div>

@@

::: {.fragment .text-xl .center}
**Think of it as your dashboard's therapist**
:::

</div>

@@

:::: {.columns}
::: {.column width="50%"}
### Before
![](img/ui-before.png)

@@

::: {.column width="50%"}
### After
![](img/ui-after.png)

>>>>

:::: {.columns}
::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="100%"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

@@

:::: {.columns}
::: {.column}
### Raise hand if you've ever:
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?

@@

::: {.column .fragment}
### 👋 Hi, I'm Jeremy!

**I'm a psychologist by training, but not the kind you're thinking of**

*Contrary to popular belief, most psychologists work with data. Not clients.*

::: {.fragment}
- **PhD in Applied Psychology**: How people actually process information
- **Led Shiny teams at scale**: *Built dashboards people use*
- **Created {bidux} package**: *Your behavioral scientist in the console*
:::
:::
::::

::: {.notes}
- 45 seconds
- Quick show of hands...who's created a user interface with more than 10 filters?
- Or if ever had a client/user ask for "just one more dropdown"
- Or built a Shiny app with more `selectInput()`s than insights
- [CLICK] If so...welcome you to my dashboard support group
- Think of our time together today as a little therapy session for dropdown overload
- [PAUSE] Hi, I'm Jeremy! I'm a psychologist by training...but not the kind you're thinking of!
- Contrary to popular belief...there are many types of psychologists...and most of us with data
- Using it to studying behavior patterns...not doing individual therapy
- PhD in Applied Psych...studied how people/groups process information and make decisions
- Transitioned from academia to industry...where I got involved leading Shiny teams at scale
- Have nearly 10 years experience building enterprise apps/dashboards folks actually use
- But it wasn't always that way...I've built MANY dashboards that flopped even though they were technically sound
    - All of these definitely apply to me
- This is what led to my core realization...
:::

## Core Realization {background-color="#5b6bbf"}

::: {.r-fit-text}
"Most dashboard failures aren't technical. They're psychological."
:::

::: {.notes}
- 15 seconds
- Most dashboard failures aren't technical, they're psychological
- That might sound obvious, but it's a critical distinction...and one that we as developers can also easily address in our workflows
- Let me show you what I mean...and how technically great products can still fail spectacularly
:::

## {background-image="img/ui-before.png" background-size="contain" background-position="right" background-color="#2c3e50"}

::: {.text-box-overlay-dark .absolute left="-150" bottom="165" .fragment}
### Real client. Real dashboard. Real problems.

@@

::: {.fragment}
**We ran telemetry for 30 days...**
:::

:::

::: {.notes}
- 45 seconds
- I worked with a client who had this 'comprehensive' dashboard
- Blinded here for confidentiality...but incredibly close to the actual design
- 18 filters, 6 tabs, every possible view, nearly zero adoption
- Why this matters: This represents thousands of developer hours and client budget wasted on technically perfect but psychologically broken software
- The client was frustrated bc they had spent months building this thing, but no one used it
- As developers, we can relate...we've all built features nobody uses
- But here's why this matters more than frustration:
    1. Technical debt from unused complexity compounds over time
    2. User abandonment patterns replicate across similar projects
    3. Without data-driven insights, we keep making the same UX mistakes
- [CLICK] So we ran telemetry for 30 days...
- This data changed everything about how we approached the dashboard design
:::

## Telemetry Unraveled the User Journey {background-color="#2c3e50"}

@@

ggplot(telemetry_data, aes(x = metric, y = percentage)) +
  geom_col(
    fill = c("#27ae60", "#e74c3c", "#c0392b", "#a93226", "#e74c3c"),
    width = 0.7
  ) +
  geom_text(
    aes(label = paste0(percentage, "%")),
    vjust = -0.5,
    size = 10,
    fontface = "bold",
    # color = "white"
  ) +
  geom_segment(
    aes(
      x = as.numeric(metric) - 0.35,
      xend = as.numeric(metric) + 0.35,
      y = percentage,
      yend = percentage
    ),
    # color = "white",
    size = 1,
    alpha = 0.3
  ) +
  labs(
    subtitle = "1,847 sessions tracked over 30 days",
    x = "",
    y = "Percentage of Users"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    # plot.background = element_rect(fill = "#e74c3c", color = NA),
    # panel.background = element_rect(fill = "#e74c3c", color = NA),
    plot.title = element_text(size = 36, face = "bold"),
    plot.subtitle = element_text(size = 24),
    axis.text.x = element_text(
      size = 16,
      angle = 25,
      hjust = 1
    ),
    axis.text.y = element_text(size = 18),
    # axis.title = element_text(color = "white"),
    panel.grid = element_blank()
  ) +
  ylim(0, 110)
```

@@

::: {.notes}
- 45 seconds
- Look at this journey of despair
- 100% open, only 35% click anything
- Only 8% succeed in finding what they needed
- And we didn't collected data on this..but I would hypothesize that many of those folks were a bit more data savvy and just knew where to look
- [CLICK] Overall...this was essentially 92% failure rate on a product that cost the client thousands of dollars to build
- This is behavioral science **screaming** at us that something is wrong here
:::

@@

@@

:::: {.columns .fragment}
### Classic Example: The Paradox of Choice

::: {.column width="50%"}
![Jam Study Results](img/jam-study-behavioral.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::

@@

::: {.notes}
- 1 minute
- Behavioral science studies how people think, feel, and act...often in the context of decision making
- "Irrationally but predictably"
- "Classic behavioral science example that perfectly mirrors our dashboard problem"
- "Sheena Iyengar's famous jam study at Columbia University"
- Show image: Side-by-side grocery store displays - one with 24 jam varieties (crowded, overwhelming), one with 6 varieties (clean, organized). Include simple bar charts showing the dramatic difference in purchase rates.
- "More choice attracted browsers, but paralyzed buyers"
- [CLICK] "This is exactly what happens with dashboard filters - more options kill engagement"
- "This principle drives everything we'll build with the BID framework"
:::

## Applied to Dashboards {background-color="#5b6bbf"}

- **Cognitive Load**: Too many choices
- **Progressive Disclosure**: Reveal complexity gradually
- **Framing Effects**: Same data, different story
- **Choice Architecture**: Guide without restricting

::: {.notes}
- 20 seconds
- "Applied to dashboards"
- "These aren't just theories, they're tools"
:::

@@

@@

@@

:::: {.columns}
::: {.column width="50%"}
### Before

![](img/ui-before.png)

@@

::: {.column width="50%"}
### After

![](img/ui-after.png)
